---
title: "Training on buccal samples"
author: "Chiara Herzog"
date: "2023-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)

# Install libraries if not available (+ load)
for (i in c('dplyr', 'ggplot2', 'here', 'fs', 'tibble', 'glmnet', 'rms', 'devtools', 'plotly')){
  if(length(find.package(i, quiet = T)) == 0){
    cat('Installing', i)
    install.packages(i); library(i, character.only = T)
  }
}

if(length(find.package('eutopsQC', quiet = T)) == 0){
    cat('Installing', i)
    devtools::install_github('chiaraherzog/eutopsQC'); library(i, character.only = T)
}

library(here)
here::i_am("2-training/3-blood/1-training.Rmd")

# Source scripts
source(here("0-data", "src", "el-classifier.R"))
source(here("0-data", "src", "train-linear-classifier.R"))

# Variables
setname = 'discovery_blood'
sampletypename = 'blood'
foldername <- '3-blood'
```


```{r}
# Load data
load(here("0-data", "data.Rdata"))
db_path <- fs::path_expand('~/Dropbox')
load(here(db_path, "data", "3c-blood", "beta_merged.Rdata"))

# Load EWAS results
db <- readRDS(here("1-ewas", foldername, "delta-beta.Rds"))

# Load mapping compatible
v1v2 <- eutopsQC::mapping_compatible_v1v2
```

# Overview

* Training is done on
  + top pval
  + top db_epi (low neutrophil)
  + top db_imm (high neutrophil)
  + combined (ranked)
  + entire beta matrix (lasso only)
  
* Each training step has lasso, ridge, elastic net (alpha = 0.5)

## Prepare db

```{r}
db <- as.data.frame(db) |> 
  tibble::rownames_to_column('cg') |> 
  dplyr::filter(cg %in% v1v2$EPICv1_Loci)
```

## Prepare pheno

```{r}
pheno_tr <- droplevels(data[data$set == setname & data$subset == 'training' & data$sampletype == sampletypename,]) |> 
  dplyr::mutate(type = factor(type, levels = c("Control", "BC case")))

pheno_val <- droplevels(data[data$set == setname & data$subset == 'internal_validation' & data$sampletype == sampletypename,]) |> 
  dplyr::mutate(type = factor(type, levels = c("Control", "BC case")))
```

# Training

```{r}
n.seq <- c(2, 5, 10, 20, 25, 50, 75, 100, 200, 500,seq(1000,30000,by=1000))
```

## P value

* Keep top 30.000

```{r}
# Input cpg list
cg_input <- db |> 
  dplyr::arrange(type) |> 
  dplyr::slice(1:30000) |> 
  dplyr::pull(cg)

# Beta subsets
beta_tr <- beta_merged[cg_input,pheno_tr$basename]
beta_val <- beta_merged[cg_input,pheno_val$basename]

if(!identical(colnames(beta_tr), pheno_tr$basename)){
  stop('names not overlapping')
}

if(!identical(colnames(beta_val), pheno_val$basename)){
  stop('names not overlapping')
}


# Train index
res_ridge <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_ridge, file = here("2-training", foldername, "res_ridge_p.Rdata"))

res_lasso <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 1,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_lasso, file = here("2-training", foldername, "res_lasso_p.Rdata"))

res_elnet <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0.5,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_elnet, file = here("2-training", foldername, "res_elnet_p.Rdata"))


rm(beta_tr, beta_val, res_lasso, res_ridge);gc()
```


## db_epi (low neutro)

```{r}
# Input cpg list
cg_input <- db |> 
  dplyr::arrange(desc(abs(db_epithelial))) |> 
  dplyr::slice(1:30000) |> 
  dplyr::pull(cg)

# Beta subsets
beta_tr <- beta_merged[cg_input,pheno_tr$basename]
beta_val <- beta_merged[cg_input,pheno_val$basename]

if(!identical(colnames(beta_tr), pheno_tr$basename)){
  stop('names not overlapping')
}

if(!identical(colnames(beta_val), pheno_val$basename)){
  stop('names not overlapping')
}


# Train index
res_ridge <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_ridge, file = here("2-training", foldername, "res_ridge_epi.Rdata"))

res_lasso <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 1,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_lasso, file = here("2-training", foldername, "res_lasso_epi.Rdata"))

res_elnet <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0.5,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_elnet, file = here("2-training", foldername, "res_elnet_epi.Rdata"))


rm(beta_tr, beta_val, res_lasso, res_ridge);gc()
```


## db_imm

```{r}
# Input cpg list
cg_input <- db |> 
  dplyr::arrange(desc(abs(db_immune))) |> 
  dplyr::slice(1:30000) |> 
  dplyr::pull(cg)

# Beta subsets
beta_tr <- beta_merged[cg_input,pheno_tr$basename]
beta_val <- beta_merged[cg_input,pheno_val$basename]

if(!identical(colnames(beta_tr), pheno_tr$basename)){
  stop('names not overlapping')
}

if(!identical(colnames(beta_val), pheno_val$basename)){
  stop('names not overlapping')
}


# Train index
res_ridge <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_ridge, file = here("2-training", foldername, "res_ridge_imm.Rdata"))

res_lasso <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 1,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_lasso, file = here("2-training", foldername, "res_lasso_imm.Rdata"))

res_elnet <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0.5,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_elnet, file = here("2-training", foldername, "res_elnet_imm.Rdata"))


rm(beta_tr, beta_val, res_lasso, res_ridge);gc()
```


## combined


```{r}
cg_input_epi <- db |> 
  dplyr::arrange(desc(abs(db_epithelial))) |> 
  dplyr::slice(1:30000) |> 
  dplyr::pull(cg)

cg_input_imm <- db |> 
  dplyr::arrange(desc(abs(db_immune))) |> 
  dplyr::slice(1:30000) |> 
  dplyr::pull(cg)
   

# no of CpGs
n <- 30000

# all CpG names
cg_names <- rep(NA, 2*n)
cg_names[seq(1,2*n,by=2)] <- cg_input_epi
cg_names[seq(2,2*n,by=2)] <- cg_input_imm
   
# Take the unique CpG names
cpg_input <- unique(cg_names)[1:30000]

# Beta subsets
beta_tr <- beta_merged[cg_input,pheno_tr$basename]
beta_val <- beta_merged[cg_input,pheno_val$basename]

if(!identical(colnames(beta_tr), pheno_tr$basename)){
  stop('names not overlapping')
}

if(!identical(colnames(beta_val), pheno_val$basename)){
  stop('names not overlapping')
}


# Train index
res_ridge <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_ridge, file = here("2-training", foldername, "res_ridge_combined.Rdata"))

res_lasso <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 1,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_lasso, file = here("2-training", foldername, "res_lasso_combined.Rdata"))

res_elnet <- train_linear_classifier(beta_tr = beta_tr,
                                     type_tr = pheno_tr$type,
                                     beta_val = beta_val,
                                     type_val = pheno_val$type,
                                     alpha = 0.5,
                                     n_seq = n.seq,
                                     lambda = "unrestricted")

save(res_elnet, file = here("2-training", foldername, "res_elnet_combined.Rdata"))

rm(beta_tr, beta_val, res_lasso, res_ridge);gc()
```

## Entire beta matrix


```{r}
beta_tr <- beta_merged[,pheno_tr$basename]
beta_val <- beta_merged[,pheno_val$basename]

if(!identical(colnames(beta_tr), pheno_tr$basename)){
  stop('names not overlapping')
}

if(!identical(colnames(beta_val), pheno_val$basename)){
  stop('names not overlapping')
}

res_lasso <- el_classifier(beta_tr = beta_tr,
                           type_tr = pheno_tr$type,
                           beta_val = beta_val,
                           type_val = pheno_val$type,
                           alpha = 1,
                           lambda = "unrestricted")

res_lasso$oob_auc
res_lasso$val_roc
save(res_lasso, file =  here("2-training", foldername, "res_lasso_entirebeta.Rdata"))
```

# Evaluate performances {.tabset .tabset-fade}

## Internal validation set AUC

```{r fig.align = "center", fig.width = 10, fig.height = 4.5}
source(here("0-data", "src", "plot_performance_comparisons.R"))

load(here("2-training", foldername, "res_ridge_epi.Rdata"))
load(here("2-training", foldername, "res_lasso_epi.Rdata"))
load(here("2-training", foldername, "res_elnet_epi.Rdata"))
ridge_epi <- combine_sets(res_ridge, "Ridge - Epithelial", type = 'all')
lasso_epi <- combine_sets(res_lasso, "Lasso - Epithelial", type = 'all')
elnet_epi <- combine_sets(res_lasso, "Elastic net - Epithelial", type = 'all')

load(here("2-training", foldername, "res_ridge_imm.Rdata"))
load(here("2-training", foldername, "res_lasso_imm.Rdata"))
load(here("2-training", foldername, "res_elnet_imm.Rdata"))
ridge_imm <- combine_sets(res_ridge, "Ridge - Immune", type = 'all')
lasso_imm <- combine_sets(res_lasso, "Lasso - Immune", type = 'all')
elnet_imm <- combine_sets(res_lasso, "Elastic net - Immune", type = 'all')

load(here("2-training", foldername, "res_ridge_combined.Rdata"))
load(here("2-training", foldername, "res_lasso_combined.Rdata"))
load(here("2-training", foldername, "res_elnet_combined.Rdata"))
ridge_combined <- combine_sets(res_ridge, "Ridge - Combined", type = 'all')
lasso_combined <- combine_sets(res_lasso, "Lasso - Combined", type = 'all')
elnet_combined <- combine_sets(res_lasso, "Elastic net - Combined", type = 'all')

load(here("2-training", foldername, "res_ridge_p.Rdata"))
load(here("2-training", foldername, "res_lasso_p.Rdata"))
load(here("2-training", foldername, "res_elnet_p.Rdata"))
ridge_p <- combine_sets(res_ridge, "Ridge - pval", type = 'all')
lasso_p <- combine_sets(res_lasso, "Lasso - pval", type = 'all')
elnet_p <- combine_sets(res_lasso, "Elastic net - pval", type = 'all')

pdat <- rbind(ridge_epi, lasso_epi, elnet_epi, ridge_imm, lasso_imm, elnet_imm, ridge_combined, lasso_combined, elnet_combined, ridge_p, lasso_p, elnet_p)
pdat <- as.data.frame(pdat)

pdat$n <- as.numeric(pdat$n)
pdat$AUC_val <- as.numeric(pdat$AUC_val)
pdat$AUC_oob <- as.numeric(pdat$AUC_oob)
pdat$AUC_tr <- as.numeric(pdat$AUC_tr)
pdat$Type <- factor(pdat$Type, levels = c("Ridge - Combined", "Lasso - Combined", "Elastic net - Combined", "Ridge - Epithelial", "Lasso - Epithelial", "Elastic net - Epithelial", "Ridge - Immune", "Lasso - Immune", "Elastic net - Immune", "Ridge - pval", "Lasso - pval", "Elastic net - pval"))

plotly::ggplotly(plot_performance_compare(pdat, pdat$AUC_val, sets = 12))

# save for plotting
save(pdat, file=here("0-data","dataframes-plotting","blood-training.Rdata"))

pdat <- pdat[order(pdat$AUC_val,decreasing = TRUE),]
head(pdat)
```

Max for Ridge pval n=19,000 (AUC_val = `r signif(pdat[which.max(pdat$AUC_val),]$AUC_val)`) 

### Out-of-bag estimate AUC

```{r fig.align = "center", fig.width = 10, fig.height = 4.5}
plotly::ggplotly(plot_performance_compare(pdat, pdat$AUC_oob, sets = 12))

pdat <- pdat[order(pdat$AUC_oob,decreasing = TRUE),]
head(pdat)
```

* max oob at n=27,000 for Ridge pval (AUC_oob = `r signif(pdat[which.max(pdat$AUC_oob),]$AUC_oob)`) 


## Slope versus AUC

```{r}
p <- pdat |> 
  ggplot(aes(x = as.numeric(Slope),
             y = AUC_oob,
             colour = Type,
             label = n)) +
  geom_point() +
  coord_cartesian(
    # xlim = c(0, 2),
                  # ylim = c(0.5, 1)
                  )
plotly::ggplotly(p)


p <- pdat |> 
  ggplot(aes(x = as.numeric(Slope),
             y = AUC_val,
             colour = Type,
             label = n)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 2),
                  ylim = c(0.5, 1))
plotly::ggplotly(p)

p <- pdat |> 
  ggplot(aes(x = as.numeric(intercept),
             y = AUC_val,
             colour = Type,
             label = n)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 2),
                  ylim = c(0.5, 1))
plotly::ggplotly(p)
```


##  Whole beta lasso: Lambda versus AUC

```{r fig.align = "center", fig.width = 7, fig.height = 3.5}
load(here("2-training", foldername, "res_lasso_entirebeta.Rdata"))
plot(res_lasso$fit.cv)
```

AUC val = `r res_lasso$val_roc`
 
## Whole beta lasso: Validation set ROC curve

```{r fig.align = "center", fig.width = 3.5, fig.height = 3.5}
plot(res_lasso$val_roc)
```

The area under the curve (internal validation set) is `r res_lasso$val_roc$auc` (OOB AUC is `r res_lasso$oob_auc`).
 

# Finalize training

<!-- * n=27,000 (highest AUC oob and high AUC_val) -->
* Choosing n = 30,000 for Ridge pval to finalise index

```{r}
# Input cpg list
cg_input <- db |> 
  dplyr::arrange(type) |> 
  dplyr::slice(1:30000) |> 
  dplyr::pull(cg)

# Pheno (entire discovery set)
pheno <- droplevels(data[data$set == setname & data$sampletype == sampletypename,]) |> 
  dplyr::mutate(type = factor(type, levels = c("Control", "BC case")))

# Beta (entire discovery set)
beta <- beta_merged[cg_input,pheno$basename]

if(!identical(colnames(beta), pheno$basename)){
  stop('names not overlapping')
}

res <- el_classifier(beta_tr = beta,
                     type_tr = pheno$type,
                     beta_val = beta,
                     type_val = pheno$type,
                     alpha = 0,
                     lambda = "unrestricted")

save(res, file = here("2-training", foldername, "res_final.Rdata"))

index_coef <- coef(res$fit.cv, s = "lambda.min")
names <- rownames(index_coef)
index_coef <- as.numeric(index_coef)
names(index_coef) <- names
index_coef <- index_coef[index_coef!=0]
index_coef <- index_coef[-1]
index_coef

hist(index_coef)
saveRDS(index_coef, file=here("2-training", foldername, "index_coef.Rds"))
```

## Scale

```{r}
index <- readRDS(here("2-training", foldername, "index_coef.Rds"))
ind <- na.omit(match(names(index), rownames(beta)))
b <- beta[ind,]

ind <- na.omit(match(rownames(b), names(index)))
w <- index[ind] 

if(!identical(names(w), rownames(b))){
  stop('***names mismatch***')
}

scale <- colSums(b*w, na.rm = TRUE)

pheno$scale <- scale

pheno |> 
  ggplot(aes(x = type,
             y = scale)) +
  geom_boxplot()

# save the output as 'scale' as this will be needed for the final index computation
saveRDS(scale, file = here("2-training", foldername, "scale.Rds"))
```