---
title: "EWAS on cervical samples"
author: "Chiara Herzog"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)

# Install libraries if not available (+ load)
for (i in c('dplyr', 'ggplot2', 'here', 'fs', 'tibble')){
  if(length(find.package(i, quiet = T)) == 0){
    cat('Installing', i)
    install.packages(i); library(i, character.only = T)
  }
}

here::i_am("1-ewas/2-cervical/1-ewas-analysis.Rmd")
load(here("0-data", "data.Rdata"))
source(here("0-data", "src", "expand_celltypes.R"))
db_path <- fs::path_expand('~/Dropbox')

setname <- 'discovery_cervical'
sampletypename <- 'cervical'
filename <- '2-cervical'
```

# Overview

An EWAS is run on the entire discovery set of cervical samples. Here, we visualise covariates and prepare a script that can be run in terminal as this is quicker.

## Visualisation of covariates

```{r}
data |> 
  dplyr::filter(set == setname) |> 
  ggplot(aes(x = type,
             y = age,
             fill = type)) +
  geom_boxplot() +
  facet_wrap(~subset) +
  ggpubr::stat_compare_means(ref.group = 'Control', label = 'p.signif')

data |> 
  dplyr::filter(set == setname) |> 
  ggplot(aes(x = type,
             y = ic,
             fill = type)) +
  geom_boxplot() +
  facet_wrap(~subset) +
  ggpubr::stat_compare_means(ref.group = 'Control', label = 'p.signif')

data |> 
  dplyr::filter(set == setname) |> 
  ggplot(aes(x = type,
             y = hepidish_CD4T,
             fill = type)) +
  geom_boxplot() +
  facet_wrap(~subset) +
  ggpubr::stat_compare_means(ref.group = 'Control', label = 'p.signif')

data |> 
  dplyr::filter(set == setname) |> 
  ggplot(aes(x = type,
             y = BMI,
             fill = type)) +
  geom_boxplot() +
  facet_wrap(~subset) +
  ggpubr::stat_compare_means(ref.group = 'Control', label = 'p.signif')

data |> 
  dplyr::filter(set == setname) |> 
  ggplot(aes(x = type,fill = menopause)) +
  geom_bar() +
  facet_wrap(~subset)

data |> 
  dplyr::filter(set == setname) |> 
  ggplot(aes(x = type,
             fill = OCP_current)) +
  geom_bar() +
  facet_wrap(~subset)

data |> 
  dplyr::filter(set == setname) |> 
  expand_celltypes() |> 
  ggplot(aes(x = type,
             y = cell_prop,
             fill = type)) +
  geom_boxplot() +
  facet_grid(celltype~subset) +
  coord_flip() +
  ggpubr::stat_compare_means(ref.group = 'Control',
                             label = 'p.signif')
```


## Prepare code for EWAS

```{r}
pheno <- droplevels(data[data$set==setname & data$sampletype==sampletypename,]) |> 
  tibble::remove_rownames() |> 
  # set rownames to be compatible with script
  tibble::column_to_rownames('basename')
save(pheno, file = here("1-ewas", "2-cervical", "pheno.Rdata"))

# Set paths
pheno <- here("1-ewas","2-cervical", "pheno.Rdata")
beta <- here(db_path, "data", "3c", "beta_merged.Rdata")
output <- here("1-ewas", "2-cervical")
script <- here("0-data", "src", "estimateDeltaBeta.R")

# Out script
out <- glue::glue("source('", script, "')
     
     estimateDeltaBeta(pheno = '", pheno, "',
     beta = '", beta, "',
     output = '", output, "',
     adjustment = c('age', 'ic'))")

# Save script
write(out, here("1-ewas", "2-cervical", "1-run-db.R"))

# Run script in terminal
```

# EWAS Results


```{r}
db <- readRDS(here("1-ewas", filename, "delta-beta.Rds"))
db <- as.data.frame(db)


db |> 
  ggplot(aes(x = type)) +
  geom_histogram(bins = 100)

db$padj <- p.adjust(db$type, method = 'BH')

db |> 
  tidyr::pivot_longer(db_epithelial:db_immune,
                      names_to = 'fraction',
                      values_to = 'db') |> 
  ggplot(aes(x = db,
             fill = fraction),
         alpha = 0.5) +
  geom_histogram(bins = 100)
```

## Significant sites

* `r sum(db$type<0.05)` significant CpGs
* `r sum(p.adjust(db$type, method = 'BH') < 0.05)` after BH adjustment
* maximum p value: `r format(max(db$type[db$padj < 0.05]), scientific = T)`