---
title: "Download additional validation data"
author: "Chiara Herzog"
date: "2023-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)

# Install libraries if not available (+ load)
for (i in c('dplyr', 'ggplot2', 'here', 'fs', 'tibble', 'glmnet', 'rms', 'devtools', 'plotly','GEOquery')){
  if(length(find.package(i, quiet = T)) == 0){
    cat('Installing', i)
    install.packages(i); library(i, character.only = T)
  }
}


library(here)
library(eutopsQC)
library(GEOquery)
library(dplyr)
library(EpiDISH)
library(ChAMP)

here::i_am("3-computation/2-computation-validation-data.Rmd")

source(here("0-data", "src", "computeIndex.R"))
source(here("0-data", "src", "adjust_age_ic.R"))

load(here("0-data", "data.Rdata"))



```

# Helper functions and settings

```{r}
# Path where data will be downloaded to, locally - change to own folder
dataDir = "C:/Users/elisa/Documents/data local" 

# Function to unzip GEO files
unpack_GEO <- function(dataDir, GSE) {
  # Set path to data
  setwd(file.path(dataDir, GSE))
  
  # Get the list of files in the directory
  files <- list.files()
  
  # Get the file that ends with "_RAW.tar"
  tarFile <- files[grepl("_RAW.tar$", files)]
  
  # Unzip the raw data
  untar(tarFile)
  
  # Get the updated list of files
  files <- list.files()
  
  # Filter out file names that end with "_RAW.tar"
  filtered_files <- files[!grepl("_RAW.tar$", files)]
  
  # Filter out file names that do not end with ".gz"
  gz_files <- filtered_files[grep("\\.gz$", filtered_files)]
  
  # Print the filtered list of .gz file names for visual check
  print(gz_files)
  
  # Iteratively unpack the files
  for (fileNr in 1:length(gz_files)) {
    gunzip(gz_files[fileNr], remove = TRUE)
  }
}
```



# Wang dataset

## Retrieve files from GEO

```{r}
GSE <- "GSE237036"

gse <- getGEO(GSE)
getGEOSuppFiles(GSE,makeDirectory = T,baseDir = file.path(dataDir)) # Get files from GEO

unpack_GEO(dataDir,GSE)

pheno <- pData(gse$GSE237036_series_matrix.txt.gz) # Pheno file
save(pheno,file=file.path(dataDir,GSE,"GSE237036_pheno.Rdata")) # Save locally 
```

## Process files

```{r}
# Create 'pheno_trimmed' dataframe for QC
pheno_trimmed <- pheno %>%
  # Selects only the case/control column and supplementary_file.1
  select("disease state:ch1", "supplementary_file.1") 

pheno_trimmed$type <- pheno_trimmed$`disease state:ch1`
pheno_trimmed <- pheno_trimmed %>%
  # Extract the desired part from 'supplementary_file.1' aka basename
  mutate(basename = sub('.*/(GSM\\d+_\\d+_R\\d+C\\d+).*', '\\1', supplementary_file.1))
pheno_trimmed$supplementary_file.1 <- NULL
pheno_trimmed$`disease state:ch1` <- NULL

# Rename case and control
pheno_trimmed <- pheno_trimmed %>%
  mutate(type = ifelse(type == 'BC', 'BC case', type),
         type = ifelse(type == 'normal', 'Control', type))

save(pheno_trimmed,file=file.path(dataDir,GSE,"GSE237036_pheno_trimmed.Rdata")) # Save locally 
```

## QC

```{r}
# Create the 'qc' directory
dir.create(file.path(dataDir,GSE, "qc"), showWarnings = FALSE)

# Set paths
phenoPath = file.path(dataDir,GSE,"GSE237036_pheno_trimmed.Rdata") # Pheno file
inputDir = file.path(dataDir,GSE)
outputDir = file.path(dataDir,GSE, "qc")

# QC
preprocessData(input=inputDir, output=outputDir, report=outputDir,pheno=phenoPath)
writeLines(capture.output(sessionInfo()), "~/Desktop/sessionInfo.txt")
```

## Cell type heterogeneity

```{r}

load(file.path(dataDir,GSE,"beta_merged.Rdata"))

# EpiDISH 
frac.m <- hepidish(beta.m = beta_merged,
                   ref1.m = centEpiFibIC.m,
                   ref2.m = centBloodSub.m,
                   h.CT.idx = 3,
                   method = 'RPC',
                   maxit = 500)


pheno_trimmed <- cbind(pheno_trimmed, frac.m)

colnames(pheno_trimmed) <- c("type", "basename","epi", "fib", "hepidish_B", "hepidish_NK","hepidish_CD4T","hepidish_CD8T","hepidish_Mono","hepidish_Neutro","hepidish_Eosino" )



pheno_trimmed$set <- "validation_blood"
pheno_trimmed$subset <- "GSE237036"

save(pheno_trimmed,file=file.path(dataDir,GSE,"GSE237036_pheno_trimmed.Rdata")) # Save locally 
```


## Add to master dataframe

```{r}
load(here("0-data", "data.Rdata")) # Load master dataframe
load(file.path(dataDir,"GSE237036","GSE237036_pheno_trimmed.Rdata")) # Load pheno

# Merge data frames by matching column names
data <- merge(data, pheno_trimmed, by = intersect(names(data), names(pheno_trimmed)), all.x = TRUE, all.y=TRUE)

```


# TCGA

# Additional breast versus normal adjacent dataset

GSE225845

## Retrieve files from GEO

```{r settings}
GSE <- "GSE225845"

gse <- getGEO(GSE)
getGEOSuppFiles(GSE,makeDirectory = T,baseDir = file.path(dataDir)) # Get files from GEO
unpack_GEO(dataDir, GSE) # Unpack 

# Pheno 
pheno <- pData(gse$GSE225845_series_matrix.txt.gz)
save(pheno,file=file.path(dataDir,GSE,"pheno_raw.Rdata")) # Save locally 
```

## Process files

```{r}
# Rename columns in the pheno dataframe
pheno$type <- NULL
pheno_trimmed <- pheno %>%
  rename(basename = 'methylation id (basenames):ch1',
    sex = 'Sex:ch1',
    type = 'sample type:ch1',
    race = 'race:ch1',
    grade = 'grade:ch1',
    er = 'er status:ch1',
    diabetes = 'diabetes:ch1',
    stage = 'ajcc stage:ch1',
    age = 'age_at_surgery:ch1'
  )

pheno_trimmed <- pheno_trimmed %>%
  select(basename, type, sex, race, grade, er, diabetes, stage, age)
rownames(pheno_trimmed) <- paste(rownames(pheno_trimmed), pheno_trimmed$basename, sep = "_")
pheno_trimmed$basename <- rownames(pheno_trimmed)
pheno_trimmed$age <- as.numeric(pheno_trimmed$age)

save(pheno_trimmed,file=file.path(dataDir,GSE,"pheno_trimmed.Rdata")) # Save locally 
```

## QC

```{r}
# Create the 'qc' directory
dir.create(file.path(dataDir,GSE, "qc"), showWarnings = FALSE)

# Set paths
phenoPath = file.path(dataDir,GSE,"pheno_trimmed.Rdata") # Pheno file
inputDir = file.path(dataDir,GSE)
outputDir = file.path(dataDir,GSE, "qc")

# QC
preprocessData(input=inputDir, output=outputDir, report=outputDir,pheno=phenoPath)
writeLines(capture.output(sessionInfo()), "~/Desktop/sessionInfo.txt")
```

## Cell type heterogeneity

```{r}

load(file.path(dataDir,"GSE225845","qc","beta_merged.Rdata"))
load(here("0-data","src","cent12CT.m.rda"))
load(file.path(dataDir,"GSE225845","pheno_trimmed.Rdata"))

# EpiDISH 
frac.m <- hepidish(beta.m = beta_merged,
                   ref1.m = centEpiFibIC.m,
                   ref2.m = cent12CT.m,
                   h.CT.idx = 3,
                   method = 'RPC',
                   maxit = 500)

pheno_trimmed <- cbind(pheno_trimmed, frac.m)
pheno_trimmed$ic <- rowSums(frac.m[, 3:ncol(frac.m)])

save(pheno_trimmed,file=file.path(dataDir,GSE,"pheno_trimmed.Rdata")) # Save locally 
```









# Blood samples in cancer discordant monozygotic twin-pairs

[GEO Accession (GSE89093)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE89093)

## Retrieve files from GEO

```{r settings}
GSE <- "GSE89093"

gse <- getGEO(GSE)
getGEOSuppFiles(GSE,makeDirectory = T,baseDir = file.path(dataDir)) # Get files from GEO
unpack_GEO(dataDir, GSE) # Unpack 

# Pheno 
pheno <- pData(gse$GSE89093_series_matrix.txt.gz)
save(pheno,file=file.path(dataDir,GSE,"pheno_raw.Rdata")) # Save locally 

```


## Process files

```{r}
# Create 'pheno_trimmed' dataframe for QC
pheno$type <- NULL
pheno_trimmed <- pheno %>%
  rename(beadchip = "beadchip and order on beadchip (sentrix id):ch1",
    type = "cancer status:ch1",
    twin_id = "twin pair id:ch1",
    cancer_type = "cancer location:ch1",
    age= "age at blood sample:ch1",
    basename= "description"
  )

pheno_trimmed <- pheno_trimmed %>%
  dplyr::select(basename, type, twin_id, cancer_type,age,beadchip)
rownames(pheno_trimmed) <- pheno_trimmed$basename

save(pheno_trimmed,file=file.path(dataDir,GSE,"pheno_trimmed.Rdata")) # Save locally 


```

## QC

```{r}

# Unnormalized beta file
beta_unnormalized <- read.table(file.path(dataDir,GSE,"GSE89093_Non-normalisedBetas_92Samples.txt"),sep='\t')
rownames <-beta_unnormalized[,1]
colnames <- beta_unnormalized[1,]
beta_unnormalized <- subset(beta_unnormalized, select = -1)
beta_unnormalized <- beta_unnormalized[-1, ]
beta_unnormalized <- as.matrix(sapply(beta_unnormalized, as.numeric))

rownames(beta_unnormalized) <- rownames[-1]
colnames(beta_unnormalized) <- colnames[-1]

# Normalization
beta <- champ.norm(beta_unnormalized, arraytype = "450k", method = "BMIQ", cores = 8)

# Save normalized beta locally
save(beta, file = file.path(dataDir,GSE,"beta.Rdata"))
```

## Cell type heterogeneity

```{r}

load(file.path(dataDir,"GSE89093","beta.Rdata"))
load(here("0-data","src","cent12CT.m.rda"))
load(file.path(dataDir,"GSE89093","pheno_trimmed.Rdata"))

# EpiDISH 
frac.m <- hepidish(beta.m = beta,
                   ref1.m = centEpiFibIC.m,
                   ref2.m = cent12CT.m,
                   h.CT.idx = 3,
                   method = 'RPC',
                   maxit = 500)

pheno_trimmed <- cbind(pheno_trimmed, frac.m)
pheno_trimmed$ic <- rowSums(frac.m[, 3:ncol(frac.m)])

save(pheno_trimmed,file=file.path(dataDir,GSE,"pheno_trimmed.Rdata")) # Save locally 
```


